RVD2: An ultra-sensitive variant detection model for low-depth targeted next-generation sequencing data

### 2013-11-28 doc/bioinformatics2013/RVD2.tex
Update manuscript.

\begin{document}
%TODO: Label all results/figures with comment indicating what experiment folder those results depend on and come from.

\begin{abstract}
Next-generation sequencing technology is increasingly being used for clinical diagnostic tests. Clinical samples are often genomically heterogeneous due to low sample purity or the presence of genetic subpopulations. However, many variant calling algorithms are optimized to call single nucleotide polymorphisms in homogeneous rather than heterogeneous samples. We present a novel variant calling algorithm that uses a hierarchical Bayesian model to estimate allele frequency and call variants in heterogeneous samples. We show that our algorithm improves upon current classifiers and has higher sensitivity and specificity over wide ranges of median read depth and a minor allele frequency. We identify five mutations in the PAXP1 gene in a matched clinical breast ductal carcinoma tumor sample; two of which are loss-of-heterozygosity events.

@@

The HCC1187 cell line is derived from  epithelial cells from primary breast tissue from a 41 y/o adult with TNM stage IIA primary ductal carcinoma. A matched normal cells were derived from lymphoblastoid cells from whole blood. Sequecing libraries were prepared according to the protocal described in the original techical report~\cite{}. The raw FASTQ read files were aligned to hg19 igenome ~\cite{} using the Isaac aligner~\cite{} to generate BAM files. We used samtools mpileup to generate pileup files using hg19 igenome ~\cite{} as reference sequence.

% TODO:cite the paper Raczy C, Petrovski R, Saunders C, Chorny I, Kruglyak S, Margulies E, Chuang HY, Kallberg M, Kumar SA, Liao A, Little KM, Stromberg M, Tanner S. Isaac: Ultra-fast whole genome secondary analysis on Illumina sequencing platforms. 2013
% TODO:cite the technical report.
% TODO: check matched cells line origin
% double checked by Yuting, didn't find the matched cells line origin
% TODO: mention tumor purity here

@@

% TODO: discuss alignment protocol
% TODO: rvd2 parameters
% TODO: chi2 test
% TODO: Pubmed for peer-reviewed paper on HCC1187 sample

@@

Considering HCC1187 sample is diploid, we did two tests based on posterior density using RVD2. First, we compared the posterior density $\mu_j^{\Delta}$ to find out if $\mu_j$ in case sample was significantly higher than that in control sample and vice verse.Threshold $\tau$ was set at $1\times10^{-3}$ to avoid error brought by computer limitation and confidence level $1-\alpha$ was set at 0.95. % The difference is constant at 1.74*10^{-4}
\begin{equation}\label{eqn:bayes_test}
	\Pr( \tilde{\mu}_j^{\text{case}} - \tilde{\mu}_j^{\text{control}} \geq \tau ) > 1-\alpha,
\end{equation}
\begin{equation}\label{eqn:bayes_test}
	\Pr( \tilde{\mu}_j^{\text{control}} - \tilde{\mu}_j^{\text{case}} \geq \tau ) > 1-\alpha,
\end{equation}

%TODO: write up the somatic test section
%TODO: specify the results from somatic test or from the $\mu_j^{\Delta}$ posterior density test.

Second, we identified the gemerline status and somatic status of HCC1187 sample by looking into posterior density of control sample and case sample independently. We can compute
$[0.0,0.2],[0.2,0.8],[0.8.1]$
confidence level $1-\alpha$ was set at 0.8

\begin{equation}\label{eqn:bayes_test}
	\Pr( \tilde{\mu}_j \in [\tau_\text{l},\tau_\text{u}] ) > 1-\alpha,
\end{equation}

@@

The original research describing this sample used Strelka to identify mutations in the same sample. They identified chr7:154760439 as variant, but did not call the other four variants. In particular strelka missed the two LOH events.

% we ran strelka and it found three positions: chr7:154754371, chr7:154758813 and chr7:154760439.

@@

\paragraph{\textbf{Samtools}}We used samtools/bcftools (v0.1.19) for variant calling. mpileup was the function to call variants and bcftools was used to save the result in standard Variant Call Format (VCF) files. In mpileup, we set the -d option sufficiently high at $10^6$ to avoid coverage cap.  mpileup was able to accept multiple replicates at one time so six bam files from each replicate group were fed into mpileup for variant calling. Read depth for each sample was kept with -D enabled, and -u was turned on to make sure the output bcf files uncompressed.

\paragraph{\textbf{GATK}}
UnifiedGenotyper is the function we used in GATK (v2.1-8) to detect mutations.We followed the general GATK recomended workflow to detect variants. Due to some format  incompatibility, we applied Picard to format read group and GATK for realignment as preprocessing. When calling variants using UnifiedGenotyper, -ploid (Number of samples in each pool $\times$ Sample Ploidy) was set at 1 because our synthetic data is haploid; we change the default setting for -dcov (default max coverage cutoff of $250\times$ ) to $10^6$ to avoid downsamplig so that the coverage was comparable to other approaches. All other parameters were set at default. Six replicates was passed to GATK for variant calling.

>>>>

\begin{document}
%TODO: Label all results/figures with comment indicating what experiment folder those results depend on and come from.
\begin{abstract}
Next-generation sequencing technology is increasingly being used for clinical diagnostic tests. Clinical samples are often genomically heterogeneous due to low sample purity or the presence of genetic subpopulations. However, many variant calling algorithms are optimized to call single nucleotide polymorphisms in homogeneous rather than heterogeneous samples. We present a novel variant calling algorithm that uses a hierarchical Bayesian model to estimate allele frequency and call variants in heterogeneous samples. We show that our algorithm improves upon current classifiers and has higher sensitivity and specificity over wide ranges of median read depth and a minor allele frequency. We identify five mutations in the PAXP1 gene in a matched clinical breast ductal carcinoma tumor sample; two of which are loss-of-heterozygosity events.

@@

The HCC1187 cell line is derived from  epithelial cells from primary breast tissue from a 41 y/o adult with TNM stage IIA primary ductal carcinoma. A matched normal cells were derived from lymphoblastoid cells from whole blood. Sequecing libraries were prepared according to the protocal described in the original techical report~\cite{}. The raw FASTQ read files were aligned to hg19 igenome ~\cite{} using the Isaac aligner to generate BAM files~\citep{raczy2013isaac}. We used samtools mpileup to generate pileup files using hg19 igenome ~\cite{} as reference sequence.~\citep{Navin:2010gu}. HCC1187 dataset is a well-recognized dataset from Illumina on which different research has been done~\citep{newman2013relative,howarth2011large,howarth2007array}.

% TODO:cite the technical report.
% TODO: check matched cells line origin
% double checked by Yuting, didn't find the matched cells line origin
% TODO: mention tumor purity here

@@

% TODO: discuss alignment protocol. Done.
% TODO: rvd2 parameters. Done.
% TODO: Pubmed for peer-reviewed paper on HCC1187 sample. Done

@@

Considering HCC1187 sample is diploid, we did two tests based on posterior density using RVD2. First, we compared the posterior density $\tilde\mu_j^{\Delta}$ to find out if $\tilde\mu_j$ in case sample was significantly higher than that in control sample and vice verse.Threshold $\tau$ was set at $1\times10^{-3}$ to avoid error brought by computer limitation; confidence level $1-\alpha$ was set at 0.95. We did $\chi^2$ test afterwards to further remove false callings.  % The difference is constant at 1.74*10^{-4}
\begin{equation}\label{eqn:bayes_test_positive}
	\Pr( \tilde{\mu}_j^{\text{case}} - \tilde{\mu}_j^{\text{control}} \geq \tau ) > 1-\alpha,
\end{equation}
\begin{equation}\label{eqn:bayes_test_negative}
	\Pr( \tilde{\mu}_j^{\text{control}} - \tilde{\mu}_j^{\text{case}} \geq \tau ) > 1-\alpha,
\end{equation}

%TODO: write up the somatic test section. Done.

Second, we identified the germline status and somatic status of HCC1187 sample by looking into posterior density of control sample and case sample independently. Basically, We computed the empirical posterior probability of $\tilde\mu_j$ across interval $[\tau_\text{l},\tau_\text{u}]$ and compare it with confidence level $1-\alpha$,
\begin{equation}\label{eqn:bayes_test_interval}
	\Pr( \tilde{\mu}_j \in [\tau_\text{l},\tau_\text{u}] ) > 1-\alpha.
\end{equation}

Confidence level $1-\alpha$ was set at 0.8 due to our lower confidence on germline and somatic status.Posterior density interval for reference, heterzygous mutation and homozygous mutation were set as $[0.0,0.2],[0.2,0.8],[0.8.1]$, respectively.

@@

The original research describing this sample used Strelka to identify mutations in the same sample. They identified chr7:154760439 as variant, but did not call the other four variants. In particular strelka missed the two LOH events. % To compare the performance of RVD2 with other algorithms, we used Strelka to identify mutationin the same sample. Strelka found three positions: chr7:154754371, chr7:154758813 and chr7:154760439.

@@

\paragraph{\textbf{Samtools}}We used samtools/bcftools (v0.1.19) for variant calling. mpileup was the function to call variants and bcftools was used to save the result in standard Variant Call Format (VCF) files. In mpileup, we set the -d option sufficiently high at $10^6$ to avoid coverage cap.  Samtools mpileup was able to accept multiple replicates at one time so six bam files from each replicate group were fed into mpileup for variant calling. Read depth for each sample was kept with -D enabled, and -u was turned on to make sure the output bcf files uncompressed.

\paragraph{\textbf{GATK}}
UnifiedGenotyper is the function we used in GATK (v2.1-8) to detect mutations.We followed the general GATK recomended workflow to detect variants. Due to some format  incompatibility, we applied Picard to format read group and GATK for realignment as preprocessing. InUnifiedGenotyper, -ploid (Number of samples in each pool $\times$ Sample Ploidy) was set at 1 because our synthetic data is haploid; we change the default setting for -dcov (default max coverage cutoff of $250\times$ ) to $10^6$ to avoid downsamplig so that the coverage was comparable to other approaches. All other parameters were set at default. Six replicates was passed to GATK for variant calling.
